|

### 2020-12-18 paper/v1/broadband.speed.covid.Rmd

<!-- capability to WFH -->
The capability to work from home has previously been studied from the
perspective of whether work tasks in a given occupation both can be and are
allowed to be performed using digital technologies independently of
location or co-location with colleagues, including supervisors
[@allen2015effective; @singh2013modeling].
However, successful telecommuting also requires that the quality and
reliability of digital services, particularly home internet connection speeds,
enable the completion of work tasks with a minimum of delay or interruption.
Prior to the pandemic, the performance of broadband services with respect to
telecommuters was never tested at scale, as working from home and connecting to
colleagues and workplace resources via the internet was the purview of a small minority of workers.
Instead, leisure use in the evening, when video streaming services are at their peak,
has been used to benchmark broadband performance and service delivery by different
Internet Service Providers (ISPs), at least in the UK [@ofcom2017].
Yet the shift towards telecommuting during various stages of lockdown around the world
has been drastic and there are speculations that post-Covid, the tendency to work from
home will be much higher than pre-Covid, raising questions around whether internet
services can accommodate the increased demand.
For example, $47$% of people in employment in the UK worked soley from home in April
$2020$, whilst the same figure only reached $5$% the year before [@ons2020;
@ons2020lm2019].
A back of the envelope calculation suggests that up to 40% of the working force
could work from home on an ongoing basis [@batty2020editorial].
Similar figures have been reported for other countries [@felstead2020homeworking].
For instance, $37$% of the European workforce worked from home in April $2020$ with
countries like Finland reaching $60$% [@eurofound2020].
In the US, almost half of the working population worked from home during the same
period because of the pandemic [@brynjolfsson2020covid],
and a recent estimate indicated that $37$% of all jobs in the US can be permanently
performed entirely from home [@NBERw26948].

<!-- ICT infrastructure and digital divides-->
None of these changes could have happened in the absence of reliable information
and communication technology (ICT) infrastructure -- both in terms of software
and hardware.
But while software innovations are easily diffused across space and
society^[See for example the huge success of videoconferencing apps such as Zoom
[@marks2020zoom].],
the same does not apply for ICT hardware infrastructure such as internet
broadband connectivity.
The literature describes first level digital divides in terms of the
availability and quality of internet connectivity, such as that manifest in
different geographies in the UK [@riddlesden2014broadband; @philip2017digital].
Second level digital divides consider the presence or lack of the necessary skills
to effectively utilise digital technologies and the internet
[@blank2014dimensions; @van2011internet].
The third level focuses on the heterogenous returns of internet usage among
different socioeconomic groups and, consequently, how digital technologies can
assist in bridging or further enhancing existing socioeconomic divides.
[@stern2009levels; @van2014digital; @van2015third].
The capability to telecommute is related to all three levels of digital divides,
but more importantly leads to differentiated outcomes regarding the economic
resilience of people and places to overcome a systemic shock such as the current
pandemic.
We identify this as a new digital divide, one that fundamentally alters the
potential returns of internet use for the user and wider community,
assumes skills or functions that are present in some occupations but not in others,
and relies upon access to high quality internet services.
**WE NEED TO CONNECT OUR FINDINGS WITH THIS AND JUSTIFY THAT THIS IS INDEED A NEW DIVIDE**
As the quality of internet infrastructure and services, as well as variation in
occupations are spatially dependent and clustered in space, our approach offers a
framework for understanding which types of places, are more likely to land on the
right side of this new digital divide.
By asking how resilient broadband speeds are as experienced in different parts of
the UK during a time of extreme demand, we interrogate which places benefit from
the greater economic resilience digital technologies can offer, not only during
the pandemic, but also into the future.

The structure of this paper is as follows.
First we review the literature on telecommuting and digital divides to better
understand the origins of the new digital divide revealed by the pandemic and its
impact on the economic resilience of different places.
We then describe our data and methodology.
Our results section first offers a classification of how internet services vary
across the UK local authorities and then assesses whether these clusters replicate
or repudiate other socio-economic and geographic patterns of economic resilience.

@@

<!-- Who works from home-->
In this analysis, the terms ‘telecommuting’ and ‘working from home’ are used
interchangeably, as most remote labour during the Covid-19 crisis was carried out
in the homes of individual employees rather than any other location [@eurofound2020].
However, it should be noted that previous research has explored how telecommuting
can occur in other places, including satellite offices or on public transport
[@felstead2012rapid; @siha2006telecommuting].
Previous research has also used a variety of definitions to measure the level of
telecommuting within different workforces, distinguishing, for example, between those
directly employed, indirectly employed, self-employed, full-time or part-time, and
those who use digital technologies to work remotely full-days or part-days
[@allen2015effective; @bailey2002review; @haddad2009examination].
No matter the definition, the option and capability to telecommute or work from home
has never been equally distributed spatially or socio-economically any more than
different industries and employment opportunities have.
For example, studies from the United States, the Netherlands, and the UK found
that telecommuters are most likely to hold professional, managerial, and technical
occupations where the workforce is better educated and wealthier, and that there is
suppressed demand among women and part-time workers
[@headicar2016move; @peters2004employees; @singh2013modeling].
Opportunities for working from home during the current pandemic have likewise not
been equally spread across the workforce.
@NBERw26948 indicated that in the US, managers, educators, as well as those
working in computer-related occupations, finance, and law can easily work from home,
and that occupations with opportunities to telecommute are associated
with higher earnings.
This is not the case for the workforce occupied in more spatially fixed occupations,
from farming, construction and manufacturing to hospitality and care services.
In the US, these occupations tend to be lower-income, non-white,
without a university degree, live in rental accommodation and lack health insurance
[@NBERw27085].
Similar trends can be observed for other countries.
For example, $75$% of workers with tertiary education worked from home in Europe
during spring $2020$, whilst only $34$% of workers with secondary education and
$14$% of those primary education did so [@eurofound2020].

## Digital divides and economic resilience {#sec:2.2}
<!-- WFH and digital divides-->
Our understanding of telecommuting as a product of enabled occupations can be
described as a manifestation of the third level digital divide, as those who are able
to use digital technologies to work from home benefit from a high rate of return on
their use of the internet in terms of autonomy, flexibility, and time saved from
commuting [@peters2004employees; @siha2006telecommuting; @singh2013modeling].
These returns have been even greater during the Covid-19 crisis,
when those with the ability to telecommute also have the ability to maintain
their employment whilst protecting their health.
However, the success of these arrangements has been dependent upon the first
level digital divide, which is associated with access and quality of internet
connectivity at a time of extremely high demand.
@SALEMINK2017360 provides a systematic review of the pre-pandemic, 
first level digital divide in infrastructure quality between urban and rural areas 
in various advanced economies.
Yet whether this variation in infrastructure quality affects the spatial footprint of
telecommuting has not previously been measured.
There are indications that those who purchase high speed connections consume 
more data of all sorts and use their connections for a variety of purposes 
[@hauge2011consumer], 
and that there is a correlation between access to internet services and a
reduction in household transport spend [@bris2017ict].
Whether the implication is more internet use and less travel 
because of increased telecommuting, these studies suggest that better internet 
services enable households to make savings and efficiencies, 
an example of the first level digital divide reinforcing the third level.
Yet did the purchase of high speed connections and increased internet access also
prepare households for long-term home-working, enforced by government restrictions?
The extreme demand during the pandemic provides a new opportunity to understand 
how infrastructure accessibility, quality, and reliability affects telecommuting, 
particularly in light of the high volumes of bandwidth-intensive video conferencing 
required in order to avoid the face-to-face contact that could increase 
the spread of infection. 
We seek to answer how internet service resilience contributes to or reduces 
economic resilience 
when the latter is dependent upon the capability to work from home, 
and thus improve our understanding of this new digital divide.
**RELATED TO YOUR POINT ABOVE, WE MAY CONCLUDE THIS IS NOT A NEW DIGITAL DIVIDE, BUT THAT IS STILL ANSWERING A USEFUL QUESTION - MIGHT END UP REWORDING AS INTERACTION BETWEEN DIGITAL DIVIDES? ET: OR REINFORCING THE 3RD LEVEL DIGITAL DIVIDE**

<!-- Covid-19 and working from home, cities, urban structure -->
The multi-layered digital divides intersect with material divides and the economic
geography of the UK.
Following the regional economic resilience literature, which
underlines the differentiated capacity of cities and regions to escape or
recover from economic crises [@martin2012regional; @kitsos2018economic],
different places have different industrial and occupational profiles,
and these affect the aggregated potential capacity of places for telecommuting.
Such profiles are associated with longstanding inequalities in the UK and their
spatial representation as a North-South divide [@martin_north_south].
Various studies have illustrated severe inequalities between the north and the
south part of the UK regarding, for example, skills and human capital,
unemployment, productivity and prosperity [@lee2014grim; @mccann2020perceptions;
@dorling2018peak].
Some scholars have even argued that the UK suffers some of the highest level of
interregional inequalities in the global north [@gal2018reducing; @mccann2016uk].
Not only are all three levels of digital divides associated to a certain extent 
with or shaped by the geography of the UK, but the intersection of the digital and
material divides affects the capacity of places to overcome
some of the economic effects of the Covid-19 pandemic.
Importantly, this is the first time that digital technologies became an essential
tool for economic resilience for such a great part of the population.

@@

<!-- clustering in general -->
The starting point of our methodological framework is cluster analysis, which can
be defined within the modern machine learning framework as an unsupervised
learning task, partitioning unlabelled observations into homogeneous groups 
known as clusters [@montero2014tsclust].
The key idea is that observations within clusters tend to be more similar than
observations between clusters.
Clustering is particularly useful for exploratory studies as it identifies
structures within the data [@aghabozorgi2015time].
Therefore, cluster analysis is a widely used family of techniques in geography
[@gordon1977classification; @everitt1974cluster].
For instance, clustering methods are the basis of *geodemographics*, a research domain 
which aims to create small area indicators or typologies of neighbourhoods 
based on various and sometimes diverse variables 
[@SINGLETON2009289; @harris2005geodemographics].
Clustering techniques have also been employed to solve *regionalisation* problems
[@niesterowicz2016].

@@

Time-series clustering methods have been developed in order to deal with clustering
problems linked to, for instance, stock or other financial data, economic,
governmental or medical data as well as machine monitoring
[@aggarwal2013time; @aggarwal2001surprising; @hyndman2015large; @WARRENLIAO20051857].
The main challenge -- and also the difference with cross-sectional clustering
problems -- is data dimensionality given the multiplicity of data points for every
individual object -- local authorities in our case -- included in the data set.
Time-series are dynamic data as the value of the observations change as
a function of time [@aghabozorgi2015time].
This high dimensionality leads to
(i) computational and algorithmic challenges regarding handling these data and
building algorithms to perform clustering over long time-series, and 
(ii) open questions regarding the choice of similarity measures in order to cluster 
similar times series objects together considering the whole length of the time-series
and overcoming issues around noise, outliers and shifts [@lin2004iterative;
@aghabozorgi2015time].

Time-series clustering methods utilising the whole length of time-series can be
grouped in three categories.
The first -- model-based approaches -- is based on recovering the underlying model
for each time-series and then applying clustering algorithms on the model parameters
of each time-series [@aghabozorgi2015time].
The main criticism is the cluster accuracy for nearby clusters [@mitsa2010temporal].
The second approach is based the formation of vectors of features derived from the
original time-series.
These new data of reduced dimensionality are then clustered using conventional
clustering algorithms.

For this paper we utilise the third category of time-series clustering methods
known as shape-based approaches.
These methods match two sperate time-series objects based on the similarity of
their shapes through the calculation of distances between the shapes, 
and are thus better equipped to capture similarities between short length
time-series [@aghabozorgi2015time].
This approach serves best this paper because (i) we identify clusters of
UK local authorities with similar temporal signatures -- i.e. shapes -- of
experienced internet speeds and (ii) the length of our time-series is short
(see the data discussion in this section).

@@

<!-- introduction of our broadband speeds data -->
To assess the quality and reliability of internet across local authorities
in the UK during the time when the population were told to work from home if at
all possible we utilise unique data comprising individual internet speed tests from
Speedchecker Ltd^[[https://www.broadbandspeedchecker.co.uk/](https://www.broadbandspeedchecker.co.uk/)].
This a private company that allows internet users to check their own broadband
upload and download speeds, and stores every speed-check with timestamp and
geolocation information.
These data have been used before to assess digital divides [@riddlesden2014broadband]
and the impact of local loop unbundling regulatory processes [@nardotto2015unbundling].
By using volunteered geographic data, we are able to assess the *experienced* 
internet speed by users, which may differ from the *advertised* maximum speeds of 
Internet Service Providers (ISPs).
We are particularly interested in upload speeds and the frequency of speed tests
over the period from March to May 2020, as government statements indicate this is
when UK workers were told to work from home if at all possible **[need ref?]**.
Average upload speeds are slower than average download speeds, at $9.3$Mb/s 
mean upload speed for the whole sample during the period of interest, 
compared to $29.6$Mb/s for download speeds, but they are also less associated 
with internet-based, high-demand, leisure activities such as video streaming.
Therefore, upload speeds are more relevant to work-related activities such as 
uploading documents or two-way audio, video, and text-based communication systems.
Meanwhile, the frequency of speed tests was important in identifying 
the temporal profile which woud give us most insight into 
experienced internet service and resilience, and provide 
an indication of the volume of experience over particular units of time.

<!-- data transformation -->
The first step in the workflow after dropping some outliers following
@riddlesden2014broadband was to transform the individual, geolocated and time-stamped 
tests to more meaningful aggregates both in terms of space and time.
The frequency of testing indicates that whilst there is an overall trend of increased
testing **or increased experience of internet service NOT SURE WHAT THIS MEAND** from March to April and then a 
slight reduction from April to May, this trend masks substantial variation 
by not only the day of the week, but also time of day, as can be seen in Figure \ref{test2020}.
Thus, a daily aggregation of upload speeds would mask the variation in experience
over the course of each weekday.
Furthermore, the importance of this variation is highlighted by a comparison 
with the same period in $2019$, as in Figure \ref{test2019}, 
when the volume of testing and thus of experience of internet service quality
was much more tied to the evening leisure peak and presumably to download speeds.
Since the increase in testing during the working day in $2020$ is an indication that 
users have a greater perception of the variable quality of internet service,
particularly during a new morning peak of testing for service reliability,
we decided to include a measure of hourly variation in our temporal profiles.
However, although this is a large data set ($241,088$ individual tests during the
working hours of study period), there are not enough observations for each
Local Authority District (LAD) and for each working hour of each working day
($631$ speed tests per LAD on average) to profile speeds at that level of detail.
Therefore, we aggregate all the speed-checks during the $13$ weeks of March to May 
inclusive for weekdays in $2020$ by each hour of the day and day of the week.
As our research aims to identify the geography of internet service resilience for
work purposes, bank holidays and the hours between midnight and 6:00 were excluded,
as well as weekend days.
The composite week time-series thus comprise $18$ hours multiplied by $5$ weekdays or
$90$ time points per series.
We also aggregate these data spatially because we could not follow individuals or 
households and connect data points.
The time-series were calculated for each of the $382$ LADs in the UK, standardised, 
and then a *k*-means partitioning around medoids clustering algorithm 
was applied using DTW.
We initially run the algorithm for $k \in \mathbb{N} \bigcap [5,15]$
and used cluster validity indices (CVIs) to pick the optimal solution of $k = 13$.
Following @sardatime the majority vote for the following CVIs was used: Silhouette 
(max), Score function (max), Calinski-Harabasz (max), Davies-Bouldin (min), Modified 
Davies-Bouldin (DB*, min), Dunn (max), COP (min).

@@

<!-- metrics -->
We then review the characteristics of each cluster, including number of LADs, 
and the temporal profile of upload speed by hour of the day 
and day of the composite week.
Since the quality and reliability of internet services vary in time and space due
to both supply and demand-side influences, we use a number of different measures
to then describe experienced upload speeds per cluster.
These include: a) mean, experienced connection speed,
b) standard deviation or the amount of fluctuation from the mean, and
c) the variation in speeds at particular times of day when working from home is
more likely to take place.
We take account of all three measurements in our descriptives statistics of
upload speeds in order to determine how resilient broadband speeds are 
as experienced in different parts of the UK during a time of extreme demand.

**More data details and some figures, descriptive stats**

<!-- link to auxillary regressions method -->
The cause of these different experiences of broadband resilience may be different in
different areas, as they may reflect either similarities in patterns of demand
or similar quality of infrastructure.
Our approach is also limited by potential endogeneity, as for example, better
quality connections with high mean speeds may enable more working from home, but
greater demand may cause slower speeds, less reliability and greater variability
of speed at different times of day or week.
Therefore, we avoid attributing any cause to our analysis of the experienced level
of quality and reliability of upload speeds.
Instead, we run an auxiliary regression in order to understand how the spatial and
temporal patterns of internet service relate to the economic geography of the UK.
We discuss how the different patterns might support or undermine efforts to work
from home and maintain safe productivity and whether they reinforce existing
spatial and social inequalities.
More specifically, we estimate the following multinomial model:

@@

The identification strategy is as follows.
Based on the outcomes of the time-series clustering, we identify $J$ distinct
and crisp clusters.
We then regress this cluster membership against a vector $X_{i}$ of
socio-economic and geographic variables, which are discussed in detail in
the relevant [section](#sec:4.2).
This analysis enables us to provide a more nuanced understanding of how
telecommuting and technology intersect at a time of extreme demand, and what
lessons this time has for a future where telecommuting is likely to remain a
common means of accessing work and broadband services, as well as infrastructure,
must be fit for purpose.

@@

The temporal profiles of the local authority clusters have been summarised in
Figures \ref{UpClusterL} and \ref{UpClusterS} and Table \ref{up.cluster.descr}.
The graphs show a composite profile of mean upload speeds per hour 
per day for each cluster, with the largest, in terms of the LAD membership, 
five clusters in Figure \ref{UpClusterL}, 
and the next six in Figure \ref{UpClusterS}.
These figures and table provide a comprehensive overview of the quality and
reliability of experienced broadband in different parts of the UK, 
the temporal clusters offering a novel approach to understanding spatial disparity.

@@

>>>>

<!-- capability to WFH -->
The capability to work from home has previously been studied from the
perspective of whether work tasks in a given occupation both can be and are
allowed to be performed using digital technologies independently of
location or co-location with colleagues, including supervisors
[@allen2015effective; @singh2013modeling].
However, successful telecommuting also requires that the quality and
reliability of digital services, particularly home internet connection speeds,
enable the completion of work tasks with a minimum of delay or interruption.
Prior to the pandemic, the performance of broadband services with respect to
telecommuters was never tested at scale, as working from home and connecting to
colleagues and workplace resources via the internet was the purview of a small minority of workers.
Instead, leisure use in the evening, when video streaming services are at their peak,
has been used to benchmark broadband performance and service delivery by different
Internet Service Providers (ISPs), at least in the UK [@ofcom2017].

Yet the shift towards telecommuting during various stages of lockdown around the world
has been drastic and there are speculations that post-Covid, the tendency to work from
home will be much higher than pre-Covid, raising questions around whether internet
services can accommodate the increased demand.
For example, $47$% of people in employment in the UK worked soley from home in April
$2020$, whilst the same figure only reached $5$% the year before [@ons2020;
@ons2020lm2019].
A back of the envelope calculation suggests that up to 40% of the working force
could work from home on an ongoing basis [@batty2020editorial].
Similar figures have been reported for other countries [@felstead2020homeworking].
For instance, $37$% of the European workforce worked from home in April $2020$ with
countries like Finland reaching $60$% [@eurofound2020].
In the US, almost half of the working population worked from home during the same
period because of the pandemic [@brynjolfsson2020covid],
and a recent estimate indicated that $37$% of all jobs in the US can be permanently
performed entirely from home [@NBERw26948].

<!-- ICT infrastructure and digital divides-->
None of these changes could have happened in the absence of reliable information
and communication technology (ICT) infrastructure -- both in terms of software
and hardware.
But while software innovations are easily diffused across space and
society^[See for example the huge success of videoconferencing apps such as Zoom
[@marks2020zoom].],
the same does not apply for ICT hardware infrastructure such as internet
broadband connectivity.
The literature describes first level digital divides in terms of the
availability and quality of internet connectivity, such as that manifest in
different geographies in the UK [@riddlesden2014broadband; @philip2017digital].
Second level digital divides consider the presence or lack of the necessary skills
to effectively utilise digital technologies and the internet
[@blank2014dimensions; @van2011internet].
The third level focuses on the heterogenous returns of internet usage among
different socioeconomic groups and, consequently, how digital technologies can
assist in bridging or further enhancing existing socioeconomic divides.
[@stern2009levels; @van2014digital; @van2015third].
The capability to telecommute is related to all three levels of digital divides,
but more importantly leads to differentiated outcomes regarding the economic
resilience of people and places to overcome a systemic shock such as the current
pandemic.
The extreme level of demand for telecommuting fundamentally alters the
potential returns of internet use for the user and wider community,
assumes skills or functions that are present in some occupations but not in others,
and relies upon access to high quality internet services.

As the quality of internet infrastructure and services, as well as variation in
occupations are spatially dependent and clustered in space, our approach offers a
framework for understanding the impact of and interactions between 
the different levels of digital division in different places 
with different characteristics.
By asking how resilient broadband speeds, and particularly upload speeds are 
as experienced in different parts of the UK during a time of extreme demand, 
we interrogate which places benefit from the greater economic resilience 
digital technologies can offer, not only during the pandemic, 
but also into the future.
The structure of this paper is as follows.
First we review the literature on telecommuting and digital divides to better
understand the structural and spatial development of these practices pre-pandemic,
and thus their importance to the economic resilience of different places.
We then describe our data and methodology.
Our results section first offers a classification of how internet services vary
across the UK local authorities and then assesses whether these clusters replicate
or repudiate other socio-economic and geographic patterns of economic resilience.

@@

<!-- Who works from home-->
In this analysis, the terms ‘telecommuting’ and ‘working from home’ are used
interchangeably, as most remote labour during the Covid-19 crisis was carried out
in the homes of individual employees rather than any other location [@eurofound2020].
However, it should be noted that previous research has explored how telecommuting
can occur in other places, including satellite offices or on public transport
[@felstead2012rapid; @siha2006telecommuting].
Previous research has also used a variety of definitions to measure the level of
telecommuting within different workforces, distinguishing, for example, between those
directly employed, indirectly employed, self-employed, full-time or part-time, and
those who use digital technologies to work remotely full-days or part-days
[@allen2015effective; @bailey2002review; @haddad2009examination].
No matter the definition, the option and capability to telecommute or work from home
has never been equally distributed spatially or socio-economically any more than
different industries and employment opportunities have.
For example, studies from the United States, the Netherlands, and the UK found
that telecommuters are most likely to hold professional, managerial, and technical
occupations where the workforce is better educated and wealthier, and that there is
suppressed demand among women and part-time workers
[@headicar2016move; @peters2004employees; @singh2013modeling].

Opportunities for working from home during the current pandemic have likewise not
been equally spread across the workforce.
@NBERw26948 indicated that in the US, managers, educators, as well as those
working in computer-related occupations, finance, and law can easily work from home,
and that occupations with opportunities to telecommute are associated
with higher earnings.
This is not the case for the workforce occupied in more spatially fixed occupations,
from farming, construction and manufacturing to hospitality and care services.
In the US, these occupations tend to be lower-income, non-white,
without a university degree, live in rental accommodation and lack health insurance
[@NBERw27085].
Similar trends can be observed for other countries.
For example, $75$% of workers with tertiary education worked from home in Europe
during spring $2020$, whilst only $34$% of workers with secondary education and
$14$% of those primary education did so [@eurofound2020].

## Digital divides and economic resilience {#sec:2.2}
<!-- WFH and digital divides-->
Our understanding of telecommuting as a product of enabled occupations can be
described as a manifestation of the third level digital divide, as those who are able
to use digital technologies to work from home benefit from a high rate of return on
their use of the internet in terms of autonomy, flexibility, and time saved from
commuting [@peters2004employees; @siha2006telecommuting; @singh2013modeling].
These returns have been even greater during the Covid-19 crisis,
when those with the ability to telecommute also have the ability to maintain
their employment whilst protecting their health.
However, the success of these arrangements has been dependent upon the first
level digital divide, which is associated with access and quality of internet
connectivity at a time of extremely high demand.
@SALEMINK2017360 provides a systematic review of the pre-pandemic, 
first level digital divide in infrastructure quality between urban and rural areas 
in various advanced economies.
Rural areas, predictably, fare worse.
Yet whether this variation in infrastructure quality affects the spatial footprint of
telecommuting has not previously been measured, in part because telecommuting has
not previously been the cause of greatest demand and pressure on internet services.

There are indications that those who purchase high speed connections consume 
more data of all sorts and use their connections for a variety of purposes 
[@hauge2011consumer], 
and that there is a correlation between access to internet services and a
reduction in household transport spend [@bris2017ict].
Whether the implication is more internet use and less travel 
because of increased telecommuting, these studies suggest that better internet 
services enable households to make savings and efficiencies, 
an example of the first level digital divide reinforcing the third level.
Yet did the purchase of high speed connections and increased internet access also
prepare households for long-term home-working, enforced by government restrictions?
The extreme demand during the pandemic provides a new opportunity to understand 
how infrastructure accessibility, quality, and reliability affects telecommuting, 
particularly in light of the high volumes of bandwidth-intensive video conferencing 
required in order to avoid the face-to-face contact that could increase 
the spread of infection. 
We seek to answer how internet service resilience contributes to or reduces 
economic resilience when the latter is dependent upon 
the capability to work from home. 
We also aim to improve our understanding of the impact 
of first level digital division on telecommuting, and whether this results in 
much more fundamental third level digital division than has previously been perceived.

<!-- Covid-19 and working from home, cities, urban structure -->
Furthermore, these multi-layered digital divides intersect with material divides 
and the economic geography of the UK.
Following the regional economic resilience literature, which
underlines the differentiated capacity of cities and regions to escape or
recover from economic crises [@martin2012regional; @kitsos2018economic],
different places have different industrial and occupational profiles,
and these affect the aggregated potential capacity of places for telecommuting.
Such profiles are associated with longstanding inequalities in the UK and their
spatial representation as a North-South divide [@martin_north_south].
Various studies have illustrated severe inequalities between the north and the
south regions of England **replace 'UK'** regarding, for example, skills and human capital,
unemployment, productivity and prosperity [@lee2014grim; @mccann2020perceptions;
@dorling2018peak].
Some scholars have even argued that the UK suffers some of the highest level of
interregional inequalities in the global north [@gal2018reducing; @mccann2016uk].
Not only are all three levels of digital divides associated to a certain extent 
with or shaped by the geography of the UK, but the intersection of the digital and
material divides affects the capacity of places to overcome
some of the economic effects of the Covid-19 pandemic.
Importantly, this is the first time that digital technologies became an essential
tool for economic resilience for such a great part of the population.

@@

##Time-Series Clustering {#sec:3.1}
<!-- clustering in general -->
The starting point of our methodological framework is cluster analysis, which can
be defined within the modern machine learning framework as an unsupervised
learning task, partitioning unlabelled observations into homogeneous groups 
known as clusters [@montero2014tsclust].
The key idea is that observations within clusters tend to be more similar than
observations between clusters.
Clustering is particularly useful for exploratory studies as it identifies
structures within the data [@aghabozorgi2015time].
Therefore, cluster analysis is a widely used family of techniques in geography
[@gordon1977classification; @everitt1974cluster].
For instance, clustering methods are the basis of *geodemographics*, a research domain 
which aims to create small area indicators or typologies of neighbourhoods 
based on various and sometimes diverse variables 
[@SINGLETON2009289; @harris2005geodemographics].
Clustering techniques have also been employed to solve *regionalisation* problems
[@niesterowicz2016].

@@

Time-series clustering methods have been developed in order to deal with clustering
problems linked to, for instance, stock or other financial data, economic,
governmental or medical data as well as machine monitoring
[@aggarwal2013time; @aggarwal2001surprising; @hyndman2015large; @WARRENLIAO20051857].
The main challenge -- and also the difference with cross-sectional clustering
problems -- is data dimensionality given the multiplicity of data points for every
individual object, local authorities in our case, included in the data set.
Time-series are dynamic data as the value of the observations change as
a function of time [@aghabozorgi2015time].
This high dimensionality leads to
(i) computational and algorithmic challenges regarding handling these data and
building algorithms to perform clustering over long time-series, and 
(ii) open questions regarding the choice of similarity measures in order to cluster 
similar times series objects together considering the whole length of the time-series
and overcoming issues around noise, outliers and shifts [@lin2004iterative;
@aghabozorgi2015time].

**REMOVE THIS PARAGRAPH - IRRELEVANT DETAIL: Time-series clustering methods utilising the whole length of time-series can be grouped in three categories. The first -- model-based approaches -- is based on recovering the underlying model for each time-series and then applying clustering algorithms on the model parameters of each time-series [@aghabozorgi2015time]. The main criticism is the cluster accuracy for nearby clusters [@mitsa2010temporal].The second approach is based the formation of vectors of features derived from the original time-series. These new data of reduced dimensionality are then clustered using conventional clustering algorithms.**

For this paper we utilise a category of time-series clustering methods
known as shape-based approaches.
These methods match two sperate time-series objects based on the similarity of
their shapes through the calculation of distances between the shapes, 
and are thus better equipped to capture similarities between short length
time-series [@aghabozorgi2015time].
This approach serves best this paper because (i) we identify clusters of
UK local authorities with similar temporal signatures -- i.e. shapes -- of
experienced internet speeds and (ii) the length of our time-series is short
(see the data discussion in this section).

@@

##Experienced Broadband Speeds {#sec:3.2}
<!-- introduction of our broadband speeds data -->
To assess the quality and reliability of internet across local authorities
in the UK during the time when the population were told to work from home if at
all possible we utilise unique data comprising individual internet speed tests from
Speedchecker Ltd^[[https://www.broadbandspeedchecker.co.uk/](https://www.broadbandspeedchecker.co.uk/)].
This is a private company that allows internet users to check their own broadband
upload and download speeds, and stores every speed-check with timestamp and
geolocation information.
These data have been used before to assess digital divides [@riddlesden2014broadband]
and the impact of local loop unbundling regulatory processes [@nardotto2015unbundling].
By using volunteered geographic data, we are able to assess the *experienced* 
internet speed by users, which may differ from the *advertised* maximum speeds of 
Internet Service Providers (ISPs).

We are particularly interested in upload speeds and the frequency of speed tests
over the period from March to May 2020, as government statements indicate this is
when UK workers were told to work from home if at all possible [@GovUK2020].
Average upload speeds are slower than average download speeds, at $9.3$Mb/s 
mean upload speed for the whole sample during the period of interest, 
compared to $29.6$Mb/s for download speeds, but they are also less associated 
with internet-based, high-demand, leisure activities such as video streaming.
Therefore, upload speeds are more relevant to work-related activities such as 
uploading documents or two-way audio, video, and text-based communication systems.
Meanwhile, the frequency of speed tests was important in identifying 
the temporal profile which would give us most insight into 
experienced internet service and resilience, and provide 
an indication of the volume of experience over particular units of time.

<!-- data transformation -->
The first step in the workflow after dropping some outliers following
@riddlesden2014broadband was to transform the individual, geolocated and time-stamped 
tests to more meaningful aggregates both in terms of space and time.
The frequency of testing indicates that whilst there is an overall trend of increased
testing from March to April and then a slight reduction from April to May, 
this trend masks substantial variation by not only the day of the week, 
but also time of day, as can be seen in Figure \ref{test2020}.
Thus, a daily aggregation of upload speeds would mask the variation in experienced
service over the course of each weekday.
Furthermore, the importance of this variation is highlighted by a comparison 
with the same period in $2019$, as in Figure \ref{test2019}, 
when the volume of testing and thus of experience of internet service quality
was much more tied to the evening leisure peak and presumably to download speeds.
Since the increase in testing during the working day in $2020$ is an indication that 
users have a greater perception of the variable quality of internet service,
particularly during a new morning peak of testing for service reliability,
we decided to include a measure of hourly variation in our temporal profiles.

However, although this is a large data set -- $241,088$ individual tests performed 
during weekday hours across the study period -- 
there are not enough observations for each Local Authority District (LAD) 
and for each working hour of each working day -- $631$ speed tests per LAD on average
-- to profile speeds at that level of detail.
Therefore, we aggregate all the speed-checks during the $13$ weeks of March to May 
inclusive for weekdays in $2020$ by each hour of the day and day of the week.
As our research aims to identify the geography of internet service resilience for
work purposes, bank holidays and the hours between midnight and 6:00 were excluded,
as well as weekend days.
The composite week time-series thus comprise $18$ hours multiplied by $5$ weekdays or
$90$ time points per series.
We also aggregate these data spatially because we could not follow individuals or 
households and connect data points.
The time-series were calculated for each of the $382$ LADs in the UK, standardised, 
and then a *k*-means partitioning around medoids clustering algorithm 
was applied using DTW.
We initially run the algorithm for $k \in \mathbb{N} \bigcap [5,15]$
and used cluster validity indices (CVIs) to pick the optimal solution of $k = 13$.
Following @sardatime the majority vote for the following CVIs was used: Silhouette 
(max), Score function (max), Calinski-Harabasz (max), Davies-Bouldin (min), Modified 
Davies-Bouldin (DB*, min), Dunn (max), COP (min).

@@

<!-- metrics -->
In [section](#sec:4.1), we review the temporal profile of upload speed 
by hour of the day and day of the composite week, as well as 
the experienced speed characteristics of each cluster.
Since the quality and reliability of internet services vary in time and space due
to both supply and demand-side influences, we use a number of different measures
to describe experienced upload speeds per cluster.
These include: a) mean, experienced connection speed,
b) standard deviation or the amount of fluctuation from the mean, and
c) the variation in speeds at particular times of day when working from home is
more likely to take place.
We take account of all three measurements in our descriptives statistics of
upload speeds in order to determine how resilient broadband speeds are 
as experienced in different parts of the UK during a time of extreme demand.

<!-- link to auxillary regressions method -->
The cause of these different experiences of broadband resilience may be different in
different areas, as they may reflect either similarities in patterns of demand
or similar quality of infrastructure.
Our approach is also limited by potential endogeneity, as for example, better
quality connections with high mean speeds may enable more working from home, but
greater demand may cause slower speeds, less reliability and greater variability
of speed at different times of day or week.
Therefore, we avoid attributing any cause to our analysis of the experienced level
of quality and reliability of upload speeds.
Instead, we run an auxiliary regression in order to understand how the spatial and
temporal patterns of internet service relate to the economic geography of the UK.
More specifically, we estimate the following multinomial model:

@@

Based on the outcomes of the time-series clustering, we identify $J$ distinct
and crisp clusters.
We then regress this cluster membership against a vector $X_{i}$ of
socio-economic and geographic variables, which are discussed in detail in
the relevant [section](#sec:4.2), in order to explore how the different patterns 
might support or undermine efforts to work from home and maintain safe productivity 
and whether they reinforce existing spatial and social inequalities.
This analysis enables us to provide a more nuanced understanding of how
telecommuting and technology intersect at a time of extreme demand, and what
lessons this time has for a future where telecommuting is likely to remain a
common means of accessing work and broadband services, as well as infrastructure,
must be fit for purpose.

@@

The temporal profiles of the local authority clusters have been summarised in
Figures \ref{UpClusterL} and \ref{UpClusterS} and Table \ref{up.cluster.descr}.
The graphs show a composite profile of mean upload speeds per hour 
per day for each cluster, with the largest, in terms of the LAD membership and
population, five clusters in Figure \ref{UpClusterL}, 
and the next six in Figure \ref{UpClusterS}.
These figures and table provide a comprehensive overview of the quality and
reliability of experienced broadband in different parts of the UK, 
the temporal clusters offering a novel approach to understanding spatial disparity.

@@

Upload speeds are not used as benchmarks in the UK, nor have been seen as integral to 
universal service, considering there has never before been such extreme demand for
telecommuting and operations such as video calls.

those in cluster $6$, including people living in Birmingham, Leeds and 
Sheffield, in twelve London Boroughs, and many suburban areas and smaller cities like
Cardiff, Oxford and Cambridge 
This lower level of reliability may be as a result of increased demand, 
e.g. for telecommuting, especially as there are few truly remote areas 
in this cluster.
